---
title: "barbieQ: An R package for analysing barcode count data from clonal tracking experiments"
output: 
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Quick start to barbieQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The barbieQ package provides a series of robust statistical tools for analysing barcode count data generated from cell *clonal tracking* (*lineage tracing*) experiments. 

In these experiments, an initial cell and its offspring collectively form a *clone* (or *lineage*). A unique DNA barcode, incorporated into the genome of an initial cell, is inherited by all its progeny within the clone. This one-to-one mapping of barcodes to clones enables  tracking of clonal behaviours. By quantifying barcode counts, researchers can measure the abundance of individual clones under various experimental conditions or perturbations. 

While existing tools for barcode count data analysis primarily rely on qualitative interpretation through visualizations, they often lack robust methods to model the sources of barcode variability. ^[[barcodetrackR](https://bioconductor.org/packages/barcodetrackR/), [CellDestiny](https://github.com/TeamPerie/CellDestiny), [genBaRcode](https://CRAN.R-project.org/package=genBaRcode)]^

To address this gap, this R software package, **barbieQ**, provides advanced statistical methods to model barcode variability. The package supports preprocessing, visualization, and statistical testing to identify barcodes with significant differences in proportions or occurrences across experimental conditions. Key functionalities include initializing data structures, filtering barcodes, and applying regression models to test for significant clonal changes.

The main functions include:

 * `createBarbieQ()`
 
 * `tagTopBarcodes()`
 
 * `plotBarcodePairCorrelation()`
 
 * `clusterCorrelatingBarcodes()`
 
 * `plotSamplePairCorrelation()`
 
 * `plotBarcodeProportion()`
 
 * `testBarcodeBias()`
 
 * `plotBarcodeBiasScatterPlot()`

# Intallation

```{r install, eval=FALSE}
# You can install the development version of barbieQ like so:
BiocManager::install("barbieQ")

## Alternatively, you can install the development version of barbieQ from GitHub
# devtools::install_github("Oshlack/barbieQ")
```

# Load Dependecy

```{r library}
suppressPackageStartupMessages({
  library(barbieQ)
  library(magrittr)
  library(tidyr)
  library(dplyr)
  library(ggplot2)
  library(circlize)
  library(logistf)
  library(igraph)
  library(data.table)
  library(ComplexHeatmap)
  library(limma)
})
```

# Simulate Data to Use

An example dataset on barcode count will be implemented from published clonal track research for better demonstrating the function in the package.

```{r data}
## Sample conditions and color palettes
sampleConditions <- data.frame(
  Treat = factor(rep(c("ctrl", "drug"), each = 6)),
  Time = rep(rep(seq_len(2), each = 3), 2)
)
conditionColor <- list(
  Treat = c(ctrl = "#999999", drug = "#112233"),
  Time = c("1" = "#778899", "2" = "#998877")
)
## Barcode count data
nbarcodes <- 50
nsamples <- 12
barcodeCount <- abs(matrix(rnorm(nbarcodes * nsamples), nbarcodes, nsamples))
rownames(barcodeCount) <- paste0("Barcode", seq_len(nbarcodes))
```

# Example

## Create barbieQ Object

Please start with creating a `barbieQ` structure by taking the barcode count matrix as input, using function `createBarbieQ()`.

By creating a `barbieQ` object, you've automatically processed a series of data transformation and saved transformed data in the `barbieQ` object, for the convenience of the following analyses.

```{r example, fig.width=8, fig.height=6}
## Passing `object`, `target` and `factorColors`
exampleBBQ <- createBarbieQ(
  object = barcodeCount, target = sampleConditions, factorColors = conditionColor
)
```

## Tag Top Contributing Barcodes

A filtering step is recommended at the beginning of analysis to filter out barcodes that consistently show low counts across the dataset. The retained barcodes are considered with essential contribution, called *"top barcodes"*. 

 - By applying the function `tagTopBarcodes()` to the `barbieQ` object, you're deciding which barcodes are *"top barcodes"* while tagging them in the object.

```{r tag top}
## Tag top Barcodes
exampleBBQ <- tagTopBarcodes(barbieQ = exampleBBQ)
```

 - Once *"top barcodes"* are decided and tagged, it's good to check their contributions before actually cutting off the *"bottom barcodes"* (that show consistent low count). 
 
     - By applying the function `plotBarcodePareto()` to the `barbieQ` object, you are visualizing the contribution of each barcode colour coded by *"top"* or *"bottom"*. (Here, the contribution refers to the average abundance of individual barcodes across samples in the dataset.)
     
     - By applying the function `plotBarcodeSankey()` to the `barbieQ` object, you are visualizing the collective contribution of the *"top"* or *"bottom"* barcode groups. 

```{r tag top v1, fig.width=8, fig.height=6}
## visualize contribution of top vs. bottom barcodes
plotBarcodePareto(barbieQ = exampleBBQ) |> plot()
```

```{r tag top v2, fig.width=4, fig.height=3}
## visualize collective contribution of top vs. bottom barcodes
plotBarcodeSankey(barbieQ = exampleBBQ) |> plot()
```

 - Once you are happy with the classification of *"top"* and *"bottom"* barcodes, you can filter out the *"bottom"* barcodes by applying function `subsetBarcodes()` based on the tagged vector.
 
```{r subset top barcodes}
exampleBBQ <- subsetBarcodes(barbieQ = exampleBBQ, retainedRows = exampleBBQ$isTop$vec)
```

## Visualize Sample Correlation

To generally understand the sources of barcode variability, you can visualize sample pair-wise correlations in a check board style by applying the function `plotSamplePairCorrelation()` to the `barbieQ()` object.

```{r sample pair cor, fig.width=5, fig.height=4}
## visualize sample pair wise correlation
plotSamplePairCorrelation(barbieQ = exampleBBQ) |> plot()
```

## Test Barcodes and Identify Significant Changes

Based on the understanding of sample conditions that likely result in the differences in barcode abundance, you can robustly test the significance of the barcode changes under a specific comparison between the sample conditions, by applying the function `testBarcodeBias()` to the `barbieQ` object. The testing results will be saved in the object, and be further visualized using functions: `plotBarcodeBiasHeatmap()` and `plotBarcodeBiasScatterPlot()`.
 
 - By specifying the parameter `method` by `diffProp` (defaulted), you are testing individual barcodes' **differential proportion** between conditions.

```{r test diffProp, fig.width=5, fig.height=4}
## test Barcode differential proportion between sample groups
exampleBBQ <- testBarcodeBias(
  barbieQ = exampleBBQ,
  sampleGroups = "Treat",
  contrastLevels = c("ctrl", "drug"),
  method = "diffProp"
)
plotBarcodeBiasHeatmap(exampleBBQ) |> plot()
plotBarcodeBiasScatterPlot(exampleBBQ) |> plot()
```

 - By specifying the parameter `method` by `diffOcc`, you are testing individual barcodes' **differential occurrence** between conditions.

```{r test diffOcc, fig.width=5, fig.height=4}
## test Barcode differential occurrence between sample groups
exampleBBQ <- testBarcodeBias(
  barbieQ = exampleBBQ,
  sampleGroups = "Treat",
  contrastLevels = c("ctrl", "drug"),
  method = "diffOcc"
)
plotBarcodeBiasHeatmap(exampleBBQ) |> plot()
plotBarcodeBiasScatterPlot(exampleBBQ) |> plot()
```

# References

We are currently writing a paper to introduce the methods and approaches implemented in this `barbieQ` package and will update with a citation once available.

# SessionInfo

```{r session info}
sessionInfo()
```
