---
title: "Barbie: An R package for analysing Barcode count data from clonal tracking experiments"
output: 
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Quick start to Barbie}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The **Barbie** package provides a suite of robust statistical tools for analysing Barcode count data generated from clonal tracking experiments. By tracking the incorporation of unique DNA Barcodes into progenitor cells and sequencing the Barcodes from progeny cells, researchers can quantify clonal populations and assess their dynamics before and after experimental perturbations. **Barbie** covers data preprocessing, statistical testing, and visualizations, offering key functions for initializing data structures, identifying dominant Barcodes, and testing differential Barcode proportions or occurrences across conditions using generalized linear regression (GLM) models.

Key functions include:

 * `createBarbie()`
 
 * `tagTopBarcodes()`
 
 * `plotBarcodePairCorrelation()`
 
 * `clusterCorrelatingBarcodes()`
 
 * `plotSamplePairCorrelation()`
 
 * `plotBarcodeProportion()`
 
 * `testBarcodeBias()`
 
 * `plotBarcodeBiasScatterPlot()`

# Intallation

```{r install, eval=FALSE}
## Install BiocManager if not already installed
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
## Load BiocManager
library(BiocManager)
## Install packages
BiocManager::install("ComplexHeatmap")
BiocManager::install("limma")
install.packages("magrittr")
install.packages("tidyr")
install.packages("dplyr")
install.packages("ggplot2")
install.packages("circlize")
install.packages("logistf")
install.packages("igraph")
install.packages("data.table")
install.packages("devtools")
library(devtools)
devtools::install()
```

# Load Dependecy

```{r library}
library(Barbie)
library(magrittr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(circlize)
library(logistf)
library(igraph)
library(data.table)
library(ComplexHeatmap)
library(limma)
```

# Simulate Data to Use

An example dataset on Barcode count will be implemented from published clonal track research for better demonstrating the function in the package.

```{r data}
## Sample conditions and color palettes
sampleConditions <- data.frame(
  Treat = factor(rep(c("ctrl", "drug"), each = 6)),
  Time = rep(rep(seq_len(2), each = 3), 2)
)
conditionColor <- list(
  Treat = c(ctrl = "#999999", drug = "#112233"),
  Time = c("1" = "#778899", "2" = "#998877")
)
## Barcode count data
nbarcodes <- 50
nsamples <- 12
barcodeCount <- abs(matrix(rnorm(nbarcodes * nsamples), nbarcodes, nsamples))
rownames(barcodeCount) <- paste0("Barcode", seq_len(nbarcodes))
```

# Example

## Create Barbie Object

```{r example, fig.width=8, fig.height=6}
## Passing `object`, `target` and `factorColors`
exampleBB <- createBarbie(
  object = barcodeCount, target = sampleConditions, factorColors = conditionColor
)
```

## Tag Top Contributing Barcodes

```{r tag top, fig.width=8, fig.height=6}
## Tag top contributing Barcodes
exampleBB <- tagTopBarcodes(Barbie = exampleBB)
## visualize top vs. bottom contribution
plotBarcodePareto(Barbie = exampleBB) |> plot()
```

```{r tag top 2, fig.width=4, fig.height=3}
plotBarcodeSankey(Barbie = exampleBB) |> plot()
```

```{r barcode pair cor, fig.width=5, fig.height=3}
## visualize Barcode pairwise correlation
plotBarcodePairCorrelation(Barbie = exampleBB) |> plot()
```

## Sample Correlation

```{r sample pair cor, fig.width=5, fig.height=4}
## visualize sample pair wise correlation
plotSamplePairCorrelation(Barbie = exampleBB) |> plot()
```

## Test on Differential Barcode Proportion

```{r test diffProp, fig.width=5, fig.height=4}
## test Barcode differential proportion between sample groups
exampleBB <- testBarcodeBias(
  Barbie = exampleBB,
  sampleGroups = "Treat",
  contrastLevels = c("ctrl", "drug"),
  method = "diffProp"
)
plotBarcodeBiasHeatmap(exampleBB) |> plot()
plotBarcodeBiasScatterPlot(exampleBB) |> plot()
```

## Test on Differential Barcode Occurrence

```{r test diffOcc, fig.width=5, fig.height=4}
## test Barcode differential occurrence between sample groups
exampleBB <- testBarcodeBias(
  Barbie = exampleBB,
  sampleGroups = "Treat",
  contrastLevels = c("ctrl", "drug"),
  method = "diffOcc"
)
plotBarcodeBiasHeatmap(exampleBB) |> plot()
plotBarcodeBiasScatterPlot(exampleBB) |> plot()
```

# References

# SessionInfo

```{r session info}
sessionInfo()
```
